The Galactic Midplane Is Not a Plane: Implications for Dynamical Analysis with {\em Gaia} Data and Beyond

### 2019-04-24 paper/ms.tex
Make figures in paper correct widths

\begin{figure*}
\plotone{fig/cartoon.pdf}
\caption{Cartoon approximation showing the effect an error in the determination of the
coordinate midplane can have on orbit integration and action estimation. The
$x$-axis shows the orbital phase and the $y$-axis the vertical height. The top
gray curve depicts an example ``true'' orbit oscillating about the true
midplane (horizontal solid gray line). Consider an observer who erroneously
assumes the midplane is located at the horizontal dashed line. Suppose this
observer measures the phase-space position of the star at two different
orbital phases (purple and green points). If the observer were then to
integrate the star's orbit using a model potential with the erroneous
midplane, they would obtain the purple and green curves for the star's orbit,
respectively. The actions estimated from these two  erroneous orbits would
subsequently differ, both from each other and from the actions estimated for
the true orbit (in the potential with the correct midplane). Hence an
incorrect midplane in the potential model assumed will induce phase-dependence
in the actions estimated for a given star in that potential.}
\label{fig:cartoon}
\end{figure*}

@@

\begin{figure*}
\plotone{fig/schmactions_one_orbit.pdf}
\caption{The artificial phase-dependence in the observed actions induced by an
error in the Galactocentric coordinate system. We consider here the thick-disk
orbit, which has actions of $(J_R, J_{\phi}, J_z) = (37.9, -1520, 7.0)\,\actunit$
and $z_{\text{max}}=850\,\pc$ (see Table~\ref{tab:orbits}). We integrate the
orbit according to the procedure laid out in Section~\ref{ssec:action_comp}.
Then, we subtract $100\,\pc$ from the $z$ value ({\em upper panels}) or the
$x$ value ({\em lower panels}) of each position in the orbit, corresponding to
an erroneous observer assuming a midplane ({\em upper}) or solar radius ({\em
lower}) that is off by $100\,\pc$. We then allow our (immortal) observer to
observe the orbit over $1\,\Gyr$ and perform the same orbit integration
procedure at each timestep, and report the values of the actions. The
computation of $J_{\phi}$ is pristine to errors in $z$, with only numerical
artifacts remaining. Only small errors are induced in $J_R$, with the middle
$90\%$ of values over the $\Gyr$ being within $\sim8\%$ of the true $J_R$. As
expected, large errors are induced in $J_z$ with a $100\,\pc$ offset in $z$,
with the middle $90\%$ of values being within $\sim43\%$ of the true $J_z$.
The $x$ offset induces uncertainties in $J_R$ and $J_{\phi}$ of $\sim21\%$ and
$\sim3\%$, respectively. A $\sim3\%$ error in $J_z$ is also induced.}
\label{fig:one_orbit_wrong_ref}
\end{figure*}

\begin{figure}
\plotone{fig/schmactions_Jz_hist.pdf}
\caption{A histogram of the observed values in $J_z$ for the thick-disk orbit
assuming a $z$ offset of $100\,\pc$. One can see that if the observed $z$
values have a bias (from e.g. an incorrectly computed midplane), then the
induced error distribution in $J_z$ is decidedly non-Gaussian. Therefore, any
sort of error propagation must take this into account. Intuition for the shape
of each panel is given in the text. We also plot one half the 95\uth{}
percentile minus the 5\uth{} percentile of each distribution as a horizontal
arrow anchored on the true $J_z$ value. We call this $\Delta J_z$ and will use
it (along with the similarly defined $\Delta J_R$ and $\Delta J_{\phi}$) to
empirically describe the error distribution. We see that $\Delta J_z$ roughly
corresponds to the distance from the true $J_z$ value to one of the modes of
the error distribution.}
\label{fig:Jz_hist}
\end{figure}

@@

\begin{figure*}
\plotone{fig/schmactions_many_orbits.pdf}
\caption{We report one half the 95\uth{} minus 5\uth{} percentile of the error
in the measured action ($\Delta J_i$) from coordinate system errors for the
thin, thick, and halo orbits (Table~\ref{tab:orbits}). See discussion in the
text for the justification in using this to measure the magnitude of the
induced error. The {\em left}, {\em center}, and {\em right} panels show the
result for $J_R$, $J_{\phi}$, and $J_z$, respectively. The {\em upper} panels
consider an offset in $z$ and the {\em lower} panels consider an offset in $x$
(equivalently, an offset in the Solar radius). In some panels we also plot as
dashed lines the epicyclic prediction of the induced action error
(Equations~\ref{eq:actions_epi_error}~and~\ref{eq:Jr_epi_dx}). In the
epicyclic approximation, a $z$ offset only induces an error in $J_z$ --- for
all three orbits the epicyclic approximation is a good description of the
$J_z$ error. An $x$ offset induces an error in $J_R$ and $J_{\phi}$. The error in
$J_R$ is somewhat well-described for the thin-disk and thick-disk orbits, and
a poor description for the halo orbit. For $J_{\phi}$, the epicyclic approximation
is not a good description for any orbit.}
\label{fig:many_orbit_wrong_ref}
\end{figure*}

\begin{figure}
\plotone{fig/schmactions_many_orbits_Jz_fun.pdf}
\caption{The fractional error in $J_z$ as a function of $J_z$ for a few
different offsets in $z$. We show this for a $z$ offset of $10, 50, 100\,\pc$
(red, cyan, green respectively). As before, the error ($\Delta J_z$) is one
half the 95\uth{} minus 5\uth{} percentile of the distribution of observed
$J_z$ values over the course of the orbit. There are large errors for
thin-disk like orbits ($J_z \sim 0.5\,\actunit$), even for a small midplane
offset of $10\,\pc$. }
\label{fig:dJz_fun_Jz}
\end{figure}

@@

\begin{figure*}
\begin{center}
\includegraphics[width=7in]{fig/midplane_fit.pdf}
\end{center}
\caption{The local midplane determined at the fiducial Solar radius
($8.2\,\kpc$) for the three FIRE galaxies \mi{}, \mf{}, and \mm{} ({\em left},
{\em center}, and {\em right} panels). All calculations are done at cosmological redshift $\z =0$. The local midplane is determined at a
position $\phi$ by taking the median height of all stars within $R=0.5\,\kpc$
and $z=1\,\kpc$ (in cylindrical coordinates). The procedure is performed again
using the new height $10$ times to converge on the local midplane height. In
order to allow for the possibility that the fiducial galactocentric coordinate
system is incorrect, we subtract the best fit $A\sin{(\phi+B)}+C$ curve from
each panel --- this figure is reproduced with the original midplane
determination (i.e. before subtracting the best fit sine curve) in
Appendix~\ref{app:lsr}. We then bootstrap resample $1000$ times on all stars
within a $2\,\kpc$ height of the fiducial midplane to determine $1\,\sigma
$ error bars, which we report as dashed lines.}
\label{fig:midplane}
\end{figure*}

\begin{figure*}
\begin{center}
\includegraphics[width=5.5in]{fig/midplane_fit_chervinsim.pdf}
\end{center}
\caption{The local midplane determined at the fiducial Solar radius
($8.2\,\kpc$) for 4 different time steps from previously run toy simulations
of a Sagittarius encounter with the Milky Way \citep{2018MNRAS.481..286L}. As
before, we have subtracted the best fit $A\sin{(\phi+B)}+C$ curve to account
for inaccuracies in the galactocentric coordinate system. Error bars are
calculated as in Figure~\ref{fig:midplane}. (\Gus{Guessing here, need to
figure out what the panels actually show:}) The left two panels show the
midplane as a function of azimuth before the first encounter. As later
encounters occur, we can see that stronger midplane variations result. In
fact, one can see that an $m=2$ mode is excited in the third panel --- this is
consistent with Figure~17 of \citet{2018MNRAS.481..286L}, which shows that an
$m=2$ mode is excited (note that $m=0$ and $m=1$ modes are stronger, but these
are removed in our sine-curve subtraction).}
\label{fig:midplane_chervin}
\end{figure*}

@@

\begin{figure*}
\begin{center}
\includegraphics[width=6.8in]{fig/range_dphi.pdf}
\end{center}
\caption{The range of midplane heights encountered as a function of angular
width. At each angle $\phi$ from Figure~\ref{fig:midplane} we consider an
angular width of $\Delta \phi$ centered on $\phi$ and report the range of
midplane heights within that width. We repeat the procedure for each $\phi$
and plot the result as translucent gray lines. We also plot the mean range as
a solid blue line and the $\pm1\sigma$ lines as dashed blue lines. The upper
$x$-axis shows the chord length from the position $\phi$ to $\pm\Delta\phi/2$.
The horizontal \thincolor{} and \thickcolor{} lines shows the midplane
variation necessary to induce a $10\%$ error in $J_z$ for the thin-disk and
thick-disk orbits, respectively. A cartoon explanation of this Figure is given
in Figure~\ref{fig:fig_to_explain}. We only consider the Solar circle to
explore the impact of azimuthal midplane variations, but similar effects are
expected for radial midplane variations.}
\label{fig:range_deltaphi}
\end{figure*}

@@

\begin{figure}
\plotone{fig/cluster_offset.pdf}
\caption{The $z$ offset necessary in order for the action error induced by our
midplane effect to be comparable to the intrinsic action dispersion from a
cluster of mass $m_c$. We compute the intrinsic dispersion from
Equation~\ref{eq:action_disp} and convert the dispersion to a $z$ offset using
the result from Figure~\ref{fig:many_orbit_wrong_ref}. We consider the result
for the thin and thick orbit (Table~\ref{tab:orbits}). Nearby open clusters
(Table~\ref{tab:real_clusters}) have orbits closest to the thin orbit.}
\label{fig:cluster_offset}
\end{figure}

@@

\begin{figure*}
\plotone{fig/Rn_vs_mc.pdf}
\caption{The dynamical neighborhood (discussed in
Section~\ref{sssec:neighborhood}) around an open cluster of mass $m_c$ for the
thin-disk, and thick-disk orbits (\thincolor{} and \thickcolor{},
respectively; see Table~\ref{tab:orbits}) and for each of the FIRE galaxies
considered here ---
\mi{}, \mf{}, and \mm{} ({\em left}, {\em center}, and {\em right} panels,
respectively). Given the $z$ offset for a cluster of mass $m_c$ computed in
Figure~\ref{fig:cluster_offset}, we convert this to a chord length using the
result from Figure~\ref{fig:range_deltaphi}. We interpret this chord length as
the ``neighborhood'' around the cluster, i.e. the distance from the cluster
one could go before the $J_z$ error induced by our midplane effect is
comparable to the intrinsic action dispersion of the cluster. When the $z$
offset is larger than the maximum range in Figure~\ref{fig:range_deltaphi}, we
report the maximum chord length ($2\times8.2\,\kpc$). We aso report the same
procedure but using the $\pm1\sigma$ lines from
Figure~\ref{fig:range_deltaphi}, which we plot here as dashed lines. For the
thick orbit (\thickcolor), the neighborhood is quite large and thus the effect
is negligible. However, for the thin orbit (\thincolor), which is closest to
the orbits of nearby open clusters (see Table~\ref{tab:real_clusters}), the
neighborhood is only a few hundred $\pc$ for low-mass clusters.}
\label{fig:Rn_mc}
\end{figure*}

@@

\begin{figure*}
\plotone{fig/orbits.pdf}
\caption{The three orbits presented in Table~\ref{tab:orbits} and considered
throughout the work. We plot the thin, thick, and halo orbits in the {\em
left}, {\em center}, and {\em right} columns, respectively. The {\em upper}
row shows a plot of $x$ vs. $y$ while the {\em lower} row shows $R$ vs. $z$.}
\label{fig:plot_orbits}
\end{figure*}

@@

\begin{figure*}
\plotone{fig/lsr.pdf}
\caption{The local standard of rest (LSR) as a function of azimuth at the
fiducial Solar circle ($R_{\odot} = 8.2\,\kpc$). No best-fit subtraction is
performed here as we did in the case of the midplane
(Section~\ref{ssec:local_midplane}).}
\end{figure*}

>>>>

\begin{figure*}
\begin{center}
\includegraphics[width=\textwidth]{fig/cartoon.pdf}
\end{center}
\caption{Cartoon approximation showing the effect an error in the determination of the
coordinate midplane can have on orbit integration and action estimation. The
$x$-axis shows the orbital phase and the $y$-axis the vertical height. The top
gray curve depicts an example ``true'' orbit oscillating about the true
midplane (horizontal solid gray line). Consider an observer who erroneously
assumes the midplane is located at the horizontal dashed line. Suppose this
observer measures the phase-space position of the star at two different
orbital phases (purple and green points). If the observer were then to
integrate the star's orbit using a model potential with the erroneous
midplane, they would obtain the purple and green curves for the star's orbit,
respectively. The actions estimated from these two  erroneous orbits would
subsequently differ, both from each other and from the actions estimated for
the true orbit (in the potential with the correct midplane). Hence an
incorrect midplane in the potential model assumed will induce phase-dependence
in the actions estimated for a given star in that potential.}
\label{fig:cartoon}
\end{figure*}

@@

\begin{figure*}
\begin{center}
\includegraphics[width=\textwidth]{fig/schmactions_one_orbit.pdf}
\end{center}
\caption{The artificial phase-dependence in the observed actions induced by an
error in the Galactocentric coordinate system. We consider here the thick-disk
orbit, which has actions of $(J_R, J_{\phi}, J_z) = (37.9, -1520, 7.0)\,\actunit$
and $z_{\text{max}}=850\,\pc$ (see Table~\ref{tab:orbits}). We integrate the
orbit according to the procedure laid out in Section~\ref{ssec:action_comp}.
Then, we subtract $100\,\pc$ from the $z$ value ({\em upper panels}) or the
$x$ value ({\em lower panels}) of each position in the orbit, corresponding to
an erroneous observer assuming a midplane ({\em upper}) or solar radius ({\em
lower}) that is off by $100\,\pc$. We then allow our (immortal) observer to
observe the orbit over $1\,\Gyr$ and perform the same orbit integration
procedure at each timestep, and report the values of the actions. The
computation of $J_{\phi}$ is pristine to errors in $z$, with only numerical
artifacts remaining. Only small errors are induced in $J_R$, with the middle
$90\%$ of values over the $\Gyr$ being within $\sim8\%$ of the true $J_R$. As
expected, large errors are induced in $J_z$ with a $100\,\pc$ offset in $z$,
with the middle $90\%$ of values being within $\sim43\%$ of the true $J_z$.
The $x$ offset induces uncertainties in $J_R$ and $J_{\phi}$ of $\sim21\%$ and
$\sim3\%$, respectively. A $\sim3\%$ error in $J_z$ is also induced.}
\label{fig:one_orbit_wrong_ref}
\end{figure*}

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{fig/schmactions_Jz_hist.pdf}
\end{center}
\caption{A histogram of the observed values in $J_z$ for the thick-disk orbit
assuming a $z$ offset of $100\,\pc$. One can see that if the observed $z$
values have a bias (from e.g. an incorrectly computed midplane), then the
induced error distribution in $J_z$ is decidedly non-Gaussian. Therefore, any
sort of error propagation must take this into account. Intuition for the shape
of each panel is given in the text. We also plot one half the 95\uth{}
percentile minus the 5\uth{} percentile of each distribution as a horizontal
arrow anchored on the true $J_z$ value. We call this $\Delta J_z$ and will use
it (along with the similarly defined $\Delta J_R$ and $\Delta J_{\phi}$) to
empirically describe the error distribution. We see that $\Delta J_z$ roughly
corresponds to the distance from the true $J_z$ value to one of the modes of
the error distribution.}
\label{fig:Jz_hist}
\end{figure}

@@

\begin{figure*}
\begin{center}
\includegraphics[width=\textwidth]{fig/schmactions_many_orbits.pdf}
\end{center}
\caption{We report one half the 95\uth{} minus 5\uth{} percentile of the error
in the measured action ($\Delta J_i$) from coordinate system errors for the
thin, thick, and halo orbits (Table~\ref{tab:orbits}). See discussion in the
text for the justification in using this to measure the magnitude of the
induced error. The {\em left}, {\em center}, and {\em right} panels show the
result for $J_R$, $J_{\phi}$, and $J_z$, respectively. The {\em upper} panels
consider an offset in $z$ and the {\em lower} panels consider an offset in $x$
(equivalently, an offset in the Solar radius). In some panels we also plot as
dashed lines the epicyclic prediction of the induced action error
(Equations~\ref{eq:actions_epi_error}~and~\ref{eq:Jr_epi_dx}). In the
epicyclic approximation, a $z$ offset only induces an error in $J_z$ --- for
all three orbits the epicyclic approximation is a good description of the
$J_z$ error. An $x$ offset induces an error in $J_R$ and $J_{\phi}$. The error in
$J_R$ is somewhat well-described for the thin-disk and thick-disk orbits, and
a poor description for the halo orbit. For $J_{\phi}$, the epicyclic approximation
is not a good description for any orbit.}
\label{fig:many_orbit_wrong_ref}
\end{figure*}

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{fig/schmactions_many_orbits_Jz_fun.pdf}
\end{center}
\caption{The fractional error in $J_z$ as a function of $J_z$ for a few
different offsets in $z$. We show this for a $z$ offset of $10, 50, 100\,\pc$
(red, cyan, green respectively). As before, the error ($\Delta J_z$) is one
half the 95\uth{} minus 5\uth{} percentile of the distribution of observed
$J_z$ values over the course of the orbit. There are large errors for
thin-disk like orbits ($J_z \sim 0.5\,\actunit$), even for a small midplane
offset of $10\,\pc$. }
\label{fig:dJz_fun_Jz}
\end{figure}

@@

\begin{figure*}
\begin{center}
\includegraphics[width=\textwidth]{fig/midplane_fit.pdf}
\end{center}
\caption{The local midplane determined at the fiducial Solar radius
($8.2\,\kpc$) for the three FIRE galaxies \mi{}, \mf{}, and \mm{} ({\em left},
{\em center}, and {\em right} panels). All calculations are done at cosmological redshift $\z =0$. The local midplane is determined at a
position $\phi$ by taking the median height of all stars within $R=0.5\,\kpc$
and $z=1\,\kpc$ (in cylindrical coordinates). The procedure is performed again
using the new height $10$ times to converge on the local midplane height. In
order to allow for the possibility that the fiducial galactocentric coordinate
system is incorrect, we subtract the best fit $A\sin{(\phi+B)}+C$ curve from
each panel --- this figure is reproduced with the original midplane
determination (i.e. before subtracting the best fit sine curve) in
Appendix~\ref{app:lsr}. We then bootstrap resample $1000$ times on all stars
within a $2\,\kpc$ height of the fiducial midplane to determine $1\,\sigma
$ error bars, which we report as dashed lines.}
\label{fig:midplane}
\end{figure*}

\begin{figure*}
\begin{center}
\includegraphics[width=342.078286667pt]{fig/midplane_fit_chervinsim.pdf}
\end{center}
\caption{The local midplane determined at the fiducial Solar radius
($8.2\,\kpc$) for 4 different time steps from previously run toy simulations
of a Sagittarius encounter with the Milky Way \citep{2018MNRAS.481..286L}. As
before, we have subtracted the best fit $A\sin{(\phi+B)}+C$ curve to account
for inaccuracies in the galactocentric coordinate system. Error bars are
calculated as in Figure~\ref{fig:midplane}. (\Gus{Guessing here, need to
figure out what the panels actually show:}) The left two panels show the
midplane as a function of azimuth before the first encounter. As later
encounters occur, we can see that stronger midplane variations result. In
fact, one can see that an $m=2$ mode is excited in the third panel --- this is
consistent with Figure~17 of \citet{2018MNRAS.481..286L}, which shows that an
$m=2$ mode is excited (note that $m=0$ and $m=1$ modes are stronger, but these
are removed in our sine-curve subtraction).}
\label{fig:midplane_chervin}
\end{figure*}

@@

\begin{figure*}
\begin{center}
\includegraphics[width=\textwidth]{fig/range_dphi.pdf}
\end{center}
\caption{The range of midplane heights encountered as a function of angular
width. At each angle $\phi$ from Figure~\ref{fig:midplane} we consider an
angular width of $\Delta \phi$ centered on $\phi$ and report the range of
midplane heights within that width. We repeat the procedure for each $\phi$
and plot the result as translucent gray lines. We also plot the mean range as
a solid blue line and the $\pm1\sigma$ lines as dashed blue lines. The upper
$x$-axis shows the chord length from the position $\phi$ to $\pm\Delta\phi/2$.
The horizontal \thincolor{} and \thickcolor{} lines shows the midplane
variation necessary to induce a $10\%$ error in $J_z$ for the thin-disk and
thick-disk orbits, respectively. A cartoon explanation of this Figure is given
in Figure~\ref{fig:fig_to_explain}. We only consider the Solar circle to
explore the impact of azimuthal midplane variations, but similar effects are
expected for radial midplane variations.}
\label{fig:range_deltaphi}
\end{figure*}

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{fig/cluster_offset.pdf}
\end{center}
\caption{The $z$ offset necessary in order for the action error induced by our
midplane effect to be comparable to the intrinsic action dispersion from a
cluster of mass $m_c$. We compute the intrinsic dispersion from
Equation~\ref{eq:action_disp} and convert the dispersion to a $z$ offset using
the result from Figure~\ref{fig:many_orbit_wrong_ref}. We consider the result
for the thin and thick orbit (Table~\ref{tab:orbits}). Nearby open clusters
(Table~\ref{tab:real_clusters}) have orbits closest to the thin orbit.}
\label{fig:cluster_offset}
\end{figure}

@@

\begin{figure*}
\begin{center}
\includegraphics[width=\textwidth]{fig/Rn_vs_mc.pdf}
\end{center}
\caption{The dynamical neighborhood (discussed in
Section~\ref{sssec:neighborhood}) around an open cluster of mass $m_c$ for the
thin-disk, and thick-disk orbits (\thincolor{} and \thickcolor{},
respectively; see Table~\ref{tab:orbits}) and for each of the FIRE galaxies
considered here ---
\mi{}, \mf{}, and \mm{} ({\em left}, {\em center}, and {\em right} panels,
respectively). Given the $z$ offset for a cluster of mass $m_c$ computed in
Figure~\ref{fig:cluster_offset}, we convert this to a chord length using the
result from Figure~\ref{fig:range_deltaphi}. We interpret this chord length as
the ``neighborhood'' around the cluster, i.e. the distance from the cluster
one could go before the $J_z$ error induced by our midplane effect is
comparable to the intrinsic action dispersion of the cluster. When the $z$
offset is larger than the maximum range in Figure~\ref{fig:range_deltaphi}, we
report the maximum chord length ($2\times8.2\,\kpc$). We aso report the same
procedure but using the $\pm1\sigma$ lines from
Figure~\ref{fig:range_deltaphi}, which we plot here as dashed lines. For the
thick orbit (\thickcolor), the neighborhood is quite large and thus the effect
is negligible. However, for the thin orbit (\thincolor), which is closest to
the orbits of nearby open clusters (see Table~\ref{tab:real_clusters}), the
neighborhood is only a few hundred $\pc$ for low-mass clusters.}
\label{fig:Rn_mc}
\end{figure*}

@@

\begin{figure*}
\begin{center}
\includegraphics[width=\textwidth]{fig/orbits.pdf}
\end{center}
\caption{The three orbits presented in Table~\ref{tab:orbits} and considered
throughout the work. We plot the thin, thick, and halo orbits in the {\em
left}, {\em center}, and {\em right} columns, respectively. The {\em upper}
row shows a plot of $x$ vs. $y$ while the {\em lower} row shows $R$ vs. $z$.}
\label{fig:plot_orbits}
\end{figure*}

@@

\begin{figure*}
\begin{center}
\includegraphics[width=\textwidth]{fig/lsr.pdf}
\end{center}
\caption{The local standard of rest (LSR) as a function of azimuth at the
fiducial Solar circle ($R_{\odot} = 8.2\,\kpc$). No best-fit subtraction is
performed here as we did in the case of the midplane
(Section~\ref{ssec:local_midplane}).}
\end{figure*}
